---
title: "Cell-Cycle Scoring and Regression"
output: html_notebook
---


```{r}
install.packages("Seurat")
library(Seurat)
library(ggplot2)
```

#importing the data, segregating the S phase and G2/M phase markers, and normalizing the matrix.
```{r}

# Read in the expression matrix The first row is a header row, the first column is rownames
exp.mat <- read.table(file = "nestorawa_forcellcycle_expressionMatrix.txt",
    header = TRUE, as.is = TRUE, row.names = 1)

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Create our Seurat object and complete the initalization steps
marrow <- CreateSeuratObject(counts = Matrix::Matrix(as.matrix(exp.mat), sparse = T))
marrow <- NormalizeData(marrow)
marrow <- FindVariableFeatures(marrow, selection.method = "vst")
marrow <- ScaleData(marrow, features = rownames(marrow))

```
#run a PCA on our object, using the variable genes we found in FindVariableFeatures()
```{r}

#we see that while most of the variance can be explained by lineage, PC8 and PC10 are split on cell-cycle genes including TOP2A and MKI67. We will attempt to regress this signal from the data, so that cell-cycle heterogeneity does not contribute to PCA or downstream analysis.
marrow <- RunPCA(marrow, features = VariableFeatures(marrow), ndims.print = 6:10, nfeatures.print = 10)

```
#Heatmap for PC_8 and PC_10
```{r}

DimHeatmap(marrow, dims = c(8, 10))

```
#Assign Cell-Cycle Scores
```{r}

#we assign each cell a score, based on its expression of G2/M and S phase markers
marrow <- CellCycleScoring(marrow, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

# view cell cycle scores and phase assignments
head(marrow[[]])

```
# Visualize the distribution of cell cycle markers across
```{r}

RidgePlot(marrow, features = c("PCNA", "TOP2A", "MCM6", "MKI67"), ncol = 2)

```
# Running a PCA on cell cycle genes reveals, unsurprisingly, that cells separate entirely by phase
```{r}

marrow <- RunPCA(marrow, features = c(s.genes, g2m.genes))
DimPlot(marrow)

```
#Save the plot using ggplot2 
```{r}

plot <- DimPlot(marrow) + theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) +
    guides(colour = guide_legend(override.aes = list(size = 10)))
ggsave(filename = "cell_cycle_vignette.jpg", height = 7, width = 12, plot = plot,
    quality = 50)

```
#Regress out cell cycle scores during data scaling
```{r}

marrow <- ScaleData(marrow, vars.to.regress = c("S.Score", "G2M.Score"), features = rownames(marrow))

```
# Now, a PCA on the variable genes no longer returns components associated with cell cycle
```{r}

marrow <- RunPCA(marrow, features = VariableFeatures(marrow), nfeatures.print = 10)

```
# When running a PCA on only cell cycle genes, cells no longer separate by cell-cycle phase
```{r}

marrow <- RunPCA(marrow, features = c(s.genes, g2m.genes))
DimPlot(marrow)

```
# Alternate Workflow
#The procedure above removes all signal associated with cell cycle
```{r}
#we suggest regressing out the difference between the G2M and S phase scores. This means that signals separating non-cycling cells and cycling cells will be maintained, but differences in cell cycle phase among proliferating cells (which are often uninteresting), will be regressed out of the data
marrow$CC.Difference <- marrow$S.Score - marrow$G2M.Score
marrow <- ScaleData(marrow, vars.to.regress = "CC.Difference", features = rownames(marrow))

```
# cell cycle effects strongly mitigated in PCA
```{r}

marrow <- RunPCA(marrow, features = VariableFeatures(marrow), nfeatures.print = 10)

```
# when running a PCA on cell cycle genes, actively proliferating cells remain distinct from G1 cells however, within actively proliferating cells, G2M and S phase cells group together
```{r}

marrow <- RunPCA(marrow, features = c(s.genes, g2m.genes))
DimPlot(marrow)

```